Joint Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2024-05-24 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn, lineno]{aastex631}

@@

% Author list
    \author[0000-0003-4568-7005]{Axel Donath}
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}
    \author[0000-0002-0905-7375]{Aneta Siemiginowska}
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}
    \author[0000-0002-3869-7996]{Vinay L.\ Kashyap} 
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}
    \author[0000-0002-0816-331X]{David A.\ van Dyk}
    \affil{Department of Mathematics, Imperial College London \\ London SW7 2AZ UK}
    \author[0000-0003-4428-7835]{Douglas Burke}
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}

@@

    \cite{Selig2015} and \cite{Pumpe2018} approach the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they are able to deconvolve \fermi observations
    into a mixture of point-like and diffuse emission components. While their method works well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it shows artifacts for smaller extended emission structures such as morphological details of extended sources in the Galactic Plane.

@@

    A patch is defined as a small, typically square, region of size $N_p \times N_p$ pixels, representing a local neighbourhood in a given image. Mathematically we can formulate the extraction of such patches as:
    %
    \begin{equation}
        \mathbf{x}_n = \mathbf{P}_n \mathbf{x},
    \end{equation}
    %
    where $\mathbf{P}_n$ represents a neighbourhood matrix to extract the $n$-th patch from the image $\mathbf{x}$ and where $\mathbf{x}_n$ is the vector of length $N_p^2$ representing the \enquote{flattened} image patch. By obtaining a large number of patches we can model the distribution of pixel intensities within the patches using a Gaussian Mixture Model (GMM). The GMM is a universal function approximator and thus can model arbitrary distributions. For now we assume that we have a pre-trained GMM which encodes the patch distribution of some reference image we have chosen. During the actual image reconstruction we aim to identify the most likely GMM component per patch, treating this as a discrete nuisance-parameter that is optimized along with the image $\mathbf{x}$.

@@

    Where $\mathcal{N}$ is the Gaussian normal distribution. If we knew the relevant component, $k_n$, for patch $\mathbf{x}_n$ we could derive the conditional log-prior distribution directly from Equation~\ref{eq:gmm}:
    %
    \begin{equation}
    \begin{split}
    \mathcal{P}_{\rm GMM}(\mathbf{x}_n|k_n) = -\frac{1}{2} \left[ \right. d \log(2\pi)
    + \log(|\boldsymbol \Sigma_{k_n}|)\\
    + (\mathbf{x}_n - \boldsymbol{\mu}_{k_n})^T \Sigma^{-1}_{k}(\mathbf{x}_n - \boldsymbol{\mu}_{k_n}) \left. \right].
    \label{eq:cond.log.prior}
    \end{split}
    \end{equation}
    %
    If we split the image $\mathbf{x}$ into multiple non-overlapping patches, for each patch $\mathbf{x}_n$, we can find the GMM component $\hat{k}_n$, which maximises the conditional log-prior distribution given in Equation~\ref{eq:cond.log.prior}:
    %
    \begin{equation}
        \hat{k}_n = \underset{k \in \{1, ..., K\} }{\argmax}{\mathcal{P}_{\rm GMM}(\mathbf{x}_n|k)}
        \label{eq:khat}
    \end{equation}
    \vspace{0.2em}
    %
    As the number of components in the GMM is typically limited to $\mathcal{O}(10^2)$ we can optimize the conditional log-prior distribution by simply evaluating it for each of the GMM components and choose the one with the maximum value. Finally, to evaluate the (optimized) log-prior of the full image we sum the maximum values of the individual patches: 
    %
    \begin{equation}
    \label{eq:prior-total}
        \mathcal{P}(\mathbf{x}, \hat{\mathbf{k}}) = 
        %\mathcal{P}(\mathbf{x} | \hat{\mathbf{k}})  + \mathcal{P}(\hat{\mathbf{k}}) =
        \sum_{n = 1}^N \mathcal{P}_{\rm GMM}(\mathbf{P}_n \mathbf{x} | \hat{k}_n) +
        \sum_{n = 1}^N \log \pi_{\hat{k}_n},
    \end{equation}   
    %
    where $\mathbf{P}_n$ is again the matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed.

@@

    A variation of cycle spinning specifically for the patch prior is proposed in \cite{Parameswaran2018}. Instead of a regular grid for the patches they propose to work with a randomised grid of patches, where the center of each patch is randomly shifted, while ensuring that patches overlap by at least one pixel. In doing so they find that a smaller total number of patches can be used, while retaining the same reconstruction quality. This improves the computational performance of the method. We also implemented the evaluation of the patch prior on a randomized patch grid for comparison and found similar results as \cite{Parameswaran2018}: the reconstruction is slightly faster with the same reconstruction quality. However for all results shown later in the manuscript we chose the simpler cycle spinning described in the previous paragraph.

@@

    To account for both systematic effects we extend the model in Equation~\ref{eq:model} for the predicted counts, by a background normalization factor $\alpha_j$ and a linear shift in position $\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})$, which is specific for each observation. This yields the following modified model:
    %
    \begin{equation}
        \label{eq:model-npred-calibration}
        \mathbf{\lambda}_j = \mathbf{E}_j \cdot  \left(\mathrm{PSF}_j \circledast \phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})\right) + \alpha_j \cdot \mathbf{B}_j
    \end{equation}
    %
    In practice one observation is used as a reference and its corresponding systematic shifts in position $\delta_{x,j}, \delta_{y,j}$ are frozen during optimization. An example of the effect of the positional shift and background normalization can be seen in Figure~\ref{fig:example-fermi-lat-non-calibrated}.

@@

    \section{Implementation}
    \subsection{\jolideco Framework}
    The goal of the reconstruction process is to optimize the a-posteriori likelihood defined by Equation~\ref{eq:total}. Given that there is a parameter for each pixel in the image, this is a high dimensional optimization problem. Differentiable programming frameworks such as PyTorch~\citep{Pytorch2019} allow for solving high dimensional optimization problems using back-propagation. They also offer a variety of optimization methods such as stochastic gradient decent (SGD). The latter is also applicable in our case, as we work with multiple observations at the same time, where a subset of the observations represents a \enquote{mini batch} in terms of ML.

@@

    \subsection{Learning GMMs}
    \label{ssec:jolideco-gmm-library}
    The patch-based image prior described in Section~\ref{sec:patch-prior} requires a pre-trained GMM. We train the GMM by choosing a set of reference images, splitting them into patches of size $8\times8$ pixels and learning a GMM with $K$ components from the whole distribution of patches. In general the strength of the GMM prior is its high \enquote{expressiveness}. This means the prior distribution contains a large variety of structural shapes and has freedom to adapt to the data in the optimization process by choosing the most likely GMM component for each patch in the image. With radio data, \cite{Bouman2016} found their results to be mostly independent of the choice of data used for learning the GMM prior.

    However, in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of prior, simply because the prior distribution is more influential when the data is less informative. Thus, there is the potential for the choice of GMM training data to be influential on results and the prior should be adapted to the particular analysis scenario in order to avoid bias. For this reason and for purposes of testing we learned and provide a selection of GMMs to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

@@

    \subsection{Optimization Strategy}
    \label{sec:opt-strategy}
    Maximising the \aposteriori likelihood defined by Equation~\ref{eq:total} with respect to the model parameters $\mathbf{x}$ is a high dimensional optimization problem. We found the convergence greatly improved when the optimizer operates on $\log{\mathbf{x}}$. This is to be expected, as the brightness in pixels of astronomical can span multiple orders of magnitudes. We used the ADAM algorithm \citep{Kingma2014}, a stochastic gradient descent method, without decay and a fixed learning rate of $\mathrm{lr}=10^{-3}$. To further improve the rate of convergence of the optimization process we applied the following strategy:

    \begin{itemize}
        \item[1.] We start from a random image where each pixel was drawn from a normal distribution with the mean corresponding to the total mean of the flux in the image and a standard deviation of 10\% of the mean. We first estimate $\mathbf{x}$ by using a uniform prior and early stopping after 50 iterations. This corresponds approximately to a reconstruction obtained with the standard RL algorithm. 
        \item[2.] We freeze the image parameters and optimize the parameters associated with the systematic error described in Equation~\ref{eq:model-npred-calibration}.
        \item[3.] Finally we simultaneously optimize
        the image $\mathbf{x}$, the GMM component indicators for each patch, $k_n$ and the parameters associated with systematic error simultaneously until convergence. Convergence is verified visually by inspecting the time series of the log posterior distribution evaluated at the iterates or the models parameters. We adjust the number of epochs (number of passes through the whole set of observations) such that the curve had flattened out for the last few hundred iterations.
    \end{itemize}

@@

        \item[\bf M1.] \textbf{Pylira}: we used the LIRA method \citep{Esch2004} via the Pylira Python wrapper \citep{Donath2022}. We relied on the default prior definition and kept the background norm fixed. In total we ran the MCMC sampler for 5000 iterations and discarded the first 500 iterations as a \enquote{burn-in} period to allow the MCMC chain to converge. We computed the mean of the posterior distribution ignoring the burn-in period.

        \item[\bf M2.] \textbf{\jolideco (Uniform, n=10)}: we used \jolideco with a uniform prior on $\mathbf{x}$ and ran the optimizer for 10 iterations. We consider this as a baseline method reference similar to RL. While the optimization procedure is different, we expect similar behaviour for the local correlations of pixels. This ad hoc approach of early stopping is often used in practice to regularise the RL output.

        \item[\bf M3.] \textbf{\jolideco (Uniform, n=1000)}:  we used \jolideco with a uniform prior on $\mathbf{x}$, but running the optimizer for 1000 iterations. This is the same as above but without early stopping. This gives rise to the issue of RL decomposing extended emission into point sources.

        \item[\bf M4.] \textbf{\jolideco (Patch, GLEAM v0.1)}: we used \jolideco with a patch prior learned on the GLEAM data, described in Section~\ref{sssec:gleam-radio-data} and the patch normalization described by Equation~\ref{eq:patch-norm}. We used the optimization strategy described in Section~\ref{sec:opt-strategy} with 2000 epochs.

        \item[\bf M5.] \textbf{\jolideco (Patch, Zoran-Weiss)}: we used \jolideco with the patch prior provided by \cite{Zoran2011}. To be consistent with how the GMM was learned we only subtracted the mean of the patch and skipped the division by the mean we introduced in Equation~\ref{eq:patch-norm}. To make the prior applicable to astronomical images, we re-scaled the image to the range $0 > \mathbf{x}_i > 1$ by dividing them by the maximum value defined by the ground truth\footnote{This is not completely fair, as it requires knowledge of the brightest expected pixel in $\mathbf{x}$. In practice this could be estimated from the brightest pixel in the data and the shape of the PSF.}.
    \end{enumerate}

@@

    Using only the data from the \enquote{XMM} scenario results in lower quality reconstructions, due to the lower angular resolution of the data. However, we find that the spiral arm structure is reconstructed at all levels of signal strength, even when barely visible by eye. The ring structure is sharpened considerably in all cases as well. The near-by point sources on the left and right of the spiral arm structure cannot be separated into individual point sources, but remain blurred along the direction of their neighbours. This a similar situation as in the lower left of the image in the A1 point source scenario, when sources are to close by.

@@

    \textbf{Zoom B} shows the region around the putative central compact object of the SNR (\cite{Vogt2018}, \cite{Hebbar2020}, but \cite{Long2020}). While the emission is barely visible in the counts image it is clearly evident above the background in the \jolideco reconstruction. The reconstruction leaves a visible extension in the point-like source. However, at this level of analysis and signal to noise, both scenarios remain plausible: a real intrinsic extension of the feature, as well as a remaining reconstructed extension in a real point source because of low signal to noise\footnote{We note that, as illustrated in Section~\ref{subsec:methods} for Scenario~A, the remaining extension of the point source after deconvolution may be smaller if the source is brighter.}

@@

    \begin{equation}
        \label{eq:approx-sigma}
        \rho(N_{\mathrm{Data}}, N_{\mathrm{Pred}}) \approx \frac{1}{\sqrt{\pi \sigma^2}}\frac{(N_{\mathrm{Data}} - N_{\mathrm{Pred}}) \circledast g_{\sigma}} {\sqrt{N_{\mathrm{Pred}}\circledast g_{\sigma}}}.
    \end{equation}

    where $g_{\sigma}$ is a Gaussian smoothing kernel of width $\sigma=5$ pixels. The result is shown in the right panel of Figure~\ref{fig:example-fermi-lat}. The image shows no systematic effects and a spatially homogeneous distribution of the residual noise. The resulting distribution of residual significances per pixel is centered at $\mu=-0.001$ with a width of $\sigma=0.991$, indicating a close agreement with a standard normal distribution, as expected for a good model of the emission. In Figure~\ref{fig:example-fermi-lat-non-calibrated} we also show the residuals per PSF class with and without the correction defined by Equation~\ref{eq:model-npred-calibration}. More details of the \fermi analysis are available at \url{https://github.com/jolideco/jolideco-fermi-examples}

>>>>

% Define document class
\documentclass[twocolumn, linenumbers]{aastex631}

@@

% Author list
    \author[0000-0003-4568-7005]{Axel Donath}
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}
    \author[0000-0002-0905-7375]{Aneta Siemiginowska}
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}
    \author[0000-0002-3869-7996]{Vinay L.\ Kashyap} 
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}
    \author[0000-0002-0816-331X]{David A.\ van Dyk}
    \affil{Department of Mathematics, Imperial College London \\ London SW7 2AZ UK}
    \author[0000-0003-4428-7835]{Douglas Burke}
    \affil{Center for Astrophysics $|$ Harvard \& Smithsonian \\ Cambridge MA 02138 USA}
    \correspondingauthor{Axel Donath}
    \email{axel.donath@cfa.harvard.edu}

@@

    \cite{Selig2015A} and \cite{Pumpe2018} approach the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they are able to deconvolve \fermi observations
    into a mixture of point-like and diffuse emission components. While their method works well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it showed artifacts around the Galactic Plane, when applied to gamma-ray data from the Fermi-LAT \cite{Selig2015B}.

@@

    A patch is defined as a small, typically square, region of size $N_p \times N_p$ pixels, representing a local neighbourhood in a given image. Mathematically we can formulate the extraction of such patches as:
    %
    \begin{equation}
        \mathbf{x}_n = \mathbf{P}_n \mathbf{x},
    \end{equation}
    %
    where $\mathbf{P}_n$ represents a neighbourhood matrix to extract the $n$-th patch from the image $\mathbf{x}$ and where $\mathbf{x}_n$ is the vector of length $N_p^2$ representing the \enquote{flattened} image patch. By obtaining a large number of patches we can model the distribution of pixel intensities within the patches using a Gaussian Mixture Model (GMM) \citep{McLachlan2004}. The GMM is a universal function approximator and thus can model arbitrary distributions. For now we assume that we have a pre-trained GMM which encodes the patch distribution of some reference image we have chosen. During the actual image reconstruction we aim to identify the most likely GMM component per patch, treating this as a discrete nuisance-parameter that is optimized along with the image $\mathbf{x}$.

@@

    Where $\mathcal{N}$ is a multivariate Gaussian normal distribution. The distribution is parameterized by the vector $\mathbf{\mu_k}$, which a dimension equivalent to the number of pixels in a patch and represents the mean of each Gaussian component. The pairwise correlation between pixels is represented by the corresponding covariance matrix $\mathbf{\Sigma}_k$, while the weight of each component is given by $\pi_k$.

    If we knew the relevant component, $k_n$, for patch $\mathbf{x}_n$ we could derive the conditional log-prior distribution directly from Equation~\ref{eq:gmm}:
    %
    \begin{equation}
    \begin{split}
    \mathcal{P}_{\rm GMM}(\mathbf{x}_n|k_n) = -\frac{1}{2} \left[ \right. d \log(2\pi)
    + \log(|\boldsymbol \Sigma_{k_n}|)\\
    + (\mathbf{x}_n - \boldsymbol{\mu}_{k_n})^T \Sigma^{-1}_{k}(\mathbf{x}_n - \boldsymbol{\mu}_{k_n}) \left. \right].
    \label{eq:cond.log.prior}
    \end{split}
    \end{equation}
    %
    If we split the image $\mathbf{x}$ into multiple non-overlapping patches, for each patch $\mathbf{x}_n$, we can find the GMM component $\hat{k}_n$, which maximises the conditional log-prior distribution given in Equation~\ref{eq:cond.log.prior}:
    %
    \begin{equation}
        \hat{k}_n = \underset{k \in \{1, ..., K\} }{\argmax}{\mathcal{P}_{\rm GMM}(\mathbf{x}_n|k)}
        \label{eq:khat}
    \end{equation}
    \vspace{0.2em}
    %
    As the number of components in the GMM is typically limited to $\mathcal{O}(10^2)$ we can optimize the conditional log-prior distribution by simply evaluating it for each of the GMM components and choose the one with the maximum value. Finally, to evaluate the (optimized) log-prior of the full image we sum the maximum values of the individual patches: 
    %
    \begin{equation}
    \label{eq:prior-total}
        \mathcal{P}(\mathbf{x}, \hat{\mathbf{k}}) = 
        %\mathcal{P}(\mathbf{x} | \hat{\mathbf{k}})  + \mathcal{P}(\hat{\mathbf{k}}) =
        \sum_{n = 1}^N \mathcal{P}_{\rm GMM}(\mathbf{P}_n \mathbf{x} | \hat{k}_n) +
        \sum_{n = 1}^N \log \pi_{\hat{k}_n},
    \end{equation}   
    %
    where $\mathbf{P}_n$ is again the matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed and $\pi_{\hat{k}_n}$ the weight of the component $\hat{k}_n$, which maximizes the log-likelihood for the n-th patch.

@@

    A variation of cycle spinning specifically for the patch prior is proposed in \cite{Parameswaran2018}. Instead of a regular grid for the patches they propose to work with a randomised grid of patches, where the center of each patch is randomly shifted, while ensuring that patches overlap by at least one pixel. In doing so they find that a smaller total number of patches can be used, while retaining the same reconstruction quality. This improves the computational performance of the method. We also implemented the evaluation of the patch prior on a randomized patch grid for comparison and found similar results as \cite{Parameswaran2018}: the reconstruction was slightly faster with the same reconstruction quality. However we found (a) practical issues with reproducibility of results when using the method on a GPU and (b) a lot of code complexity to handle the edge of images correctly. Give the performance improvement was only small,
    we decided to just stick with the much simpler regular grid and cycle spinning, described in the previous paragraph.

@@

    To account for both systematic effects we extend the model in Equation~\ref{eq:model} for the predicted counts, by a background normalization factor $\alpha_j$ and a linear shift in position $\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})$, which is specific for each observation. This yields the following modified model:
    %
    \begin{equation}
        \label{eq:model-npred-calibration}
        \boldsymbol{\lambda}_j = \mathbf{E}_j \cdot  \left(\mathrm{PSF}_j \circledast \phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})\right) + \alpha_j \cdot \mathbf{B}_j
    \end{equation}
    %
    In practice one observation is used as a reference and its corresponding systematic shifts in position $\delta_{x,j}, \delta_{y,j}$ are frozen during optimization. An example of the effect of the positional shift and background normalization can be seen in Figure~\ref{fig:example-fermi-lat-non-calibrated}.

@@

    \section{Implementation}
    \subsection{\jolideco Framework}
    The goal of the reconstruction process is to find the mode of the posterior distribution defined by Equation~\ref{eq:total}. Given that there is a parameter for each pixel in the image, this is a high dimensional optimization problem. Differentiable programming frameworks such as PyTorch~\citep{Pytorch2019} allow for solving high dimensional optimization problems using back-propagation. They also offer a variety of optimization methods such as stochastic gradient decent (SGD). The latter is also applicable in our case, as we work with multiple observations at the same time, where a subset of the observations represents a \enquote{mini batch} in terms of ML.

@@

    \subsection{Learning GMMs}
    \label{ssec:jolideco-gmm-library}
    The patch-based image prior described in Section~\ref{sec:patch-prior} requires a pre-trained GMM. We train the GMM by choosing a set of reference images, splitting them into patches of size $8\times8$ pixels and learning a GMM with $K=128$ components from the whole distribution of patches. In general the strength of the GMM prior is its high \enquote{expressiveness}. This means the prior distribution contains a large variety of structural shapes and has freedom to adapt to the data in the optimization process by choosing the most likely GMM component for each patch in the image. \cite{Zoran2012} empirically found that the reconstruction quality depended on the number of components of the GMM. A higher number of components clearly lead to an improved reconstruction quality, as the GMM was more expressive. This also aligns with the intuitive argument that GMMs are universal function approximators; with a sufficient number of components they can approximate any distribution. \cite{Zoran2012} found the reconstruction quality started to saturate once the number of components was large enough. Empirically they measured a number of 32 components to be sufficient in practice for natural images. We decided to adopt a more conservative number of 128.

    With radio data, \cite{Bouman2016} found their results to be mostly independent of the choice of data used for learning the GMM prior. However, in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of prior, simply because the prior distribution is more influential when the data is less informative. Thus, there is the potential for the choice of GMM training data to be influential on results and the prior should be adapted to the particular analysis scenario in order to avoid bias. For this reason and for purposes of testing we learned and provide a selection of GMMs to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

@@

    \subsection{Optimization Strategy}
    \label{sec:opt-strategy}
    Maximising the \aposteriori likelihood defined by Equation~\ref{eq:total} with respect to the model parameters $\mathbf{x}$ is a high dimensional optimization problem. We found the convergence greatly improved when the optimizer operates on $\log{\mathbf{x}}$. This is to be expected, as the brightness in pixels of astronomical can span multiple orders of magnitudes. We used the ADAM algorithm \citep{Kingma2014}, a stochastic gradient descent method, without decay and a fixed learning rate of $\mathrm{lr}=10^{-3}$. In practice we found the convergence of the method to be very stable, however we applied the following strategy to speed up the optimization process:

    \begin{itemize}
        \item[1.] We start from a random image where each pixel was drawn from a normal distribution with the mean corresponding to the total mean of the flux in the image and a standard deviation of 10\% of the mean. We first estimate $\mathbf{x}$ by using a uniform prior and early stopping after 50 iterations. This corresponds approximately to a reconstruction obtained with the standard RL algorithm. As the uniform prior is much faster to evaluate ($\mathcal{O(100)}$), we quickly arrive at a much better starting estimate for the reconstructed image. 
        \item[2.] We freeze the image parameters and optimize the parameters associated with the systematic error described in Equation~\ref{eq:model-npred-calibration}. This way we arrive at a better starting estimate for those. Again the process is fast, because the number of parameters is very small. 
        \item[3.] Finally we simultaneously optimize
        the image $\mathbf{x}$, the GMM component indicators for each patch, $k_n$ and the parameters associated with systematic error simultaneously until convergence. Convergence is verified visually by inspecting the time series of the log posterior distribution evaluated at the iterates or the models parameters. We adjust the number of epochs (number of passes through the whole set of observations) such that the curve had flattened out for the last few hundred iterations.
    \end{itemize}

@@

        \item[\bf M1.] \textbf{\jolideco (Uniform, n=10)}: we used \jolideco with a uniform prior on $\mathbf{x}$ and ran the optimizer for 10 iterations. We consider this as a baseline method reference similar to RL. While the optimization procedure is different, we expect similar behaviour for the local correlations of pixels. This ad hoc approach of early stopping is often used in practice to regularise the RL output.

        \item[\bf M2.] \textbf{\jolideco (Uniform, n=1000)}:  we used \jolideco with a uniform prior on $\mathbf{x}$, but running the optimizer for 1000 iterations. This is the same as above but without early stopping. This gives rise to the issue of RL decomposing extended emission into point sources.

        \item[\bf M3.] \textbf{Pylira}: we used the LIRA method \citep{Esch2004} via the Pylira Python wrapper \citep{Donath2022}. We relied on the default prior definition and kept the background norm fixed. In total we ran the MCMC sampler for 5000 iterations and discarded the first 500 iterations as a \enquote{burn-in} period to allow the MCMC chain to converge. We computed the mean of the posterior distribution ignoring the burn-in period.

        \item[\bf M4.] \textbf{\jolideco (Patch, Zoran-Weiss)}: we used \jolideco with the patch prior provided by \cite{Zoran2011}. To be consistent with how the GMM was learned we only subtracted the mean of the patch and skipped the division by the mean we introduced in Equation~\ref{eq:patch-norm}. To make the prior applicable to astronomical images, we re-scaled the image to the range $0 > \mathbf{x}_i > 1$ by dividing them by the maximum value defined by the ground truth\footnote{This is not completely fair, as it requires knowledge of the brightest expected pixel in $\mathbf{x}$. In practice this could be estimated from the brightest pixel in the data and the shape of the PSF.}.

        \item[\bf M5.] \textbf{\jolideco (Patch, GLEAM v0.1)}: we used \jolideco with a patch prior learned on the GLEAM data, described in Section~\ref{sssec:gleam-radio-data} and the patch normalization described by Equation~\ref{eq:patch-norm}. We used the optimization strategy described in Section~\ref{sec:opt-strategy} with 2000 epochs.

    \end{enumerate}

@@

    Using only the data from the \enquote{XMM} scenario results in lower quality reconstructions, due to the lower angular resolution of the data. However, we find that the spiral arm structure is reconstructed at all levels of signal strength, even when barely visible by eye. The ring structure is sharpened considerably in all cases as well. The near-by point sources on the left and right of the spiral arm structure cannot be separated into individual point sources, but remain blurred along the direction of their neighbours. Thids a similar situation as in the lower left of the image in the A1 point source scenario, when sources are to close by.

@@

    \textbf{Zoom B} shows the region around the putative central compact object of the SNR (\cite{Vogt2018}, \cite{Hebbar2020}, but \cite{Long2020}). While the emission is barely visible in the counts image it is clearly evident above the background in the \jolideco reconstruction. Visually the reconstruction does not result in a point source, but leaves a visible extension of a very few pixels.

    However, at this level of analysis and signal to noise, both scenarios remain plausible: a real intrinsic extension of the feature, as well as a remaining reconstructed extension in a real point source because of low signal to noise, which has also been observed on the simulated point source dataset discussed in Section~\ref{subsec:test-datasets}\footnote{We note that, as illustrated in Section~\ref{subsec:methods} for Scenario~A, the remaining extension of the point source after deconvolution may be smaller if the source is brighter.}.

@@

    \begin{equation}
        \label{eq:approx-sigma}
        \rho(\mathbf{D}, \boldsymbol{\lambda}) \approx \frac{1}{\sqrt{\pi \sigma^2}}\frac{(\mathbf{D} - \boldsymbol{\lambda}) \circledast g_{\sigma}} {\sqrt{\boldsymbol{\lambda}\circledast g_{\sigma}}}.
    \end{equation}

    where $\mathbf{D}$ is the measured counts image, $\boldsymbol{\lambda}$ the predicted counts image  and $g_{\sigma}$ is a Gaussian smoothing kernel of width $\sigma=5$ pixels. The Gaussian smoothing kernel approximates the shape and size of the \fermi PSF. Together with the definition above, the standardized residuals approximately represent the significance of detecting a point source at the given pixel. Structures smaller than the PSF cannot be real.

    The result is shown in the right panel of Figure~\ref{fig:example-fermi-lat}. The image shows no systematic effects and a spatially homogeneous distribution of the residual noise. The resulting distribution of residual significances per pixel is centered at $\mu=-0.001$ with a width of $\sigma=0.991$, indicating a close agreement with a standard normal distribution, as expected for a good model of the emission. In Figure~\ref{fig:example-fermi-lat-non-calibrated} we also show the residuals per PSF class with and without the correction defined by Equation~\ref{eq:model-npred-calibration}. More details of the \fermi analysis are available at \url{https://github.com/jolideco/jolideco-fermi-examples}
